---
title: "march_madness_2021"
author: "Jared Klug"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(caret)
library(corrplot)
library(glmnet)
```

# Introduction

In this project, we have gotten data from [Kaggle](https://www.kaggle.com/c/ncaam-march-mania-2021), which contains extensive data of regular season games, and conferences for the NCAA Basketball seasons from 2003-2021. Using this data, I would like to predict the potential scores of the upcoming games in the 2021 March Madness Tournament

```{r gather data, message = F, warning = F}
og_df =read.csv("./data/ncaam-march-mania-2021/MRegularSeasonDetailedResults.csv") %>% 
  rename_at(vars(starts_with("W")), ~str_replace(., "W", "T1_")) %>% 
  rename_at(vars(starts_with("L")), ~str_replace(., "L", "T2_")) %>% 
  mutate(
    point_diff = T1_Score - T2_Score
  ) %>% 
  select(-Season, -DayNum , -T1_TeamID, -T1_Score, -T2_TeamID, -T2_Score, -T1_Loc, -NumOT)



#Game-by-Game Data
df = read.csv("./data/ncaam-march-mania-2021/MRegularSeasonDetailedResults.csv") %>%
  rowid_to_column("game_id") %>%
   relocate(WLoc:NumOT, .after = DayNum) %>%
  mutate(
    Wpoint_diff = WScore - LScore,
    Lpoint_diff = LScore - WScore
  )%>%
  pivot_longer(
    WTeamID:Lpoint_diff,
    names_to = "stat",
    values_to = "count"
  ) %>%
  mutate(outcome = case_when(
    str_detect(stat, "^W") ~ "win",
    str_detect(stat, "^L") ~ "loss"
  )) %>%
  mutate(stat = substr(stat, 2, nchar(stat))) %>%
  pivot_wider(
    names_from = stat,
    values_from = count
  ) %>%
  mutate(TeamID = as.factor(TeamID)) %>%
  unnest()
  
team_id = read.csv("./data/ncaam-march-mania-2021/MTeams.csv") %>% 
  select(TeamID, TeamName)

df[["TeamID"]] = team_id[match(df[["TeamID"]], team_id[["TeamID"]]) , "TeamName"]

df = df %>% 
  rename(
    TeamName = TeamID
  )


#Season Data
season_df = df %>% 
  group_by(Season, TeamName) %>% 
  summarize_at(vars(Score:point_diff), ~mean(.x))

#2020 Avg Season Data for each team (used as input for predictions)

MM21_teams = c("Gonzaga", "Baylor", "Illinois", "Michigan", "Alabama", "Ohio St", "Iowa", "Houston", 
               "Arkansas", "West Virginia", "Texas", "Kansas", "Florida St", "Purdue", "Oklahoma St", "Virginia",
               "Creighton", "Villanova", "Tennessee", "Colorado", "USC", "Texas Tech", "BYU", "San Diego St", 
               "Oregon", "Connecticut", "Clemson", "Florida", "LSU", "Loyola-Chicago", "North Carolina", "Oklahoma",
               "Missouri", "Georgia Tech", "Wisconsin", "Maryland", "St Bonaventure", "Virginia Tech", "VCU", "Rutgers",
               "Syracuse", "Utah St", "Michigan St", "UCLA", "Wichita St", "Oregon St", "Georgetown", "Drake",
               "Winthrop", "UC Santa Barbara", "Ohio", "North Texas", "Liberty", "UNC Greensboro", "Abilene Chr", "Morehead St",
               "Colgate", "E Washington", "Grand Canyon", "Cleveland St", "Oral Roberts", "Iona", "Drexel", "Hartford",
               "Mt St Mary's", "TX Southern", "Norfolk St", "Appalachian St")

season_20_tstat = season_df %>% 
  filter(Season == 2020,
         TeamName %in% MM21_teams) %>% 
  ungroup() %>% 
  select(-Season, -Score, -point_diff)


```

```{r}
set.seed(2021)

og_df2 = model.matrix(point_diff ~ ., og_df)[, -1]

trainRows = createDataPartition(y = og_df$point_diff, p = 0.8, list = F)

x = og_df2[trainRows,]

y = og_df$point_diff[trainRows]


ctrl1 = trainControl(method = "repeatedcv", number = 10, repeats = 5)

enet.fit = train(x,y,
                 method = "glmnet",
                 tuneGrid = expand.grid(alpha = seq(0, 1, length = 15),
                                         lambda = exp(seq(1, -5, length = 50))),
                 trControl = ctrl1)

enet.fit$bestTune

coef(enet.fit$finalModel, enet.fit$bestTune$lambda)

enet.pred = predict(enet.fit, newdata = og_df2[-trainRows,])

mean((enet.pred - og_df$point_diff[-trainRows])^2)


```

```{r}
#Generate 2021 Match-ups (assuming every team will play each other)
matches = t(combn(MM21_teams, 2)) %>% 
  as.data.frame() %>% 
  rename(
    T1_name = V1,
    T2_name = V2
  )

matches = left_join(matches, season_20_tstat, by = c("T1_name" = "TeamName"))

colnames(matches)[3:15] = paste("T1_", colnames(matches)[3:15], sep = "")

matches = left_join(matches, season_20_tstat, by = c("T2_name" = "TeamName"))
  
colnames(matches)[16:28] = paste("T2_", colnames(matches)[16:28], sep = "")

MM21_pred = predict(enet.fit, newdata = matches)

test1 = cbind(matches[1:2], MM21_pred)

find_matchup = function(t1_name, t2_name){
  test = try(filter(test1, 
             T1_name == t1_name,
             T2_name == t2_name), silent = T)
  
   if(nrow(test) == 0) {
     test = try(filter(test1, 
             T2_name == t1_name,
             T1_name == t2_name), silent = T)
   }
  return(test)
}
```

# Using Four Factors

Shooting - Effective Field Goal Percentage (eFG%) = (FGM + 0.5 * FGM3) / FGA

Turnovers - Turnover Percentage (TOV%) = (TOV / (FGA + 0.44 * FTA + TOV)

Rebounding - Rebound Percentage (OR% and DR%):
  OR% = OR / (OR + Opp DR)
  DR% = DR / (DR + Opp OR)Free Throws = FTM / FGA
  
Free Throws - Free Throw Percentage (FT%) = FTA / FTM
```{r}
ff_df = read.csv("./data/ncaam-march-mania-2021/MRegularSeasonDetailedResults.csv") %>% 
  rename_at(vars(starts_with("W")), ~str_replace(., "W", "T1_")) %>% 
  rename_at(vars(starts_with("L")), ~str_replace(., "L", "T2_")) %>% 
  mutate(
    point_diff = T1_Score - T2_Score
  )

ff_df[["T1_TeamID"]] = team_id[match(ff_df[["T1_TeamID"]], team_id[["TeamID"]]) , "TeamName"]
ff_df[["T2_TeamID"]] = team_id[match(ff_df[["T2_TeamID"]], team_id[["TeamID"]]) , "TeamName"]

get_summed_stats = function(teamName){
  
  df1 = ff_df %>% 
    filter(T1_TeamID == teamName) %>% 
    group_by(Season) %>% 
    summarize(
      sum_o.fgm = sum(T1_FGM),
      sum_o.fgm3 = sum(T1_FGM3),
      sum_o.fga = sum(T1_FGA, T1_FGA3),
      sum_o.or = sum(T1_OR),
      sum_o.df = sum(T1_DR),
      sum_o.to = sum(T1_TO),
      sum_o.ftm = sum(T1_FTM),
      sum_o.fta = sum(T1_FTA),
      sum_d.fgm = sum(T2_FGM),
      sum_d.fgm3 = sum(T2_FGM3),
      sum_d.fga = sum(T2_FGA, T2_FGA3),
      sum_d.or = sum(T2_OR),
      sum_d.df = sum(T2_DR),
      sum_d.to = sum(T2_TO),
      sum_d.ftm = sum(T2_FTM),
      sum_d.fta = sum(T2_FTA)      
  )
  
  df2 = ff_df %>% 
    filter(T2_TeamID == teamName) %>% 
    group_by(Season) %>% 
    summarize(
      sum_o.fgm = sum(T2_FGM),
      sum_o.fgm3 = sum(T2_FGM3),
      sum_o.fga = sum(T2_FGA, T2_FGA3),
      sum_o.or = sum(T2_OR),
      sum_o.df = sum(T2_DR),
      sum_o.to = sum(T2_TO),
      sum_o.ftm = sum(T2_FTM),
      sum_o.fta = sum(T2_FTA),
      sum_d.fgm = sum(T1_FGM),
      sum_d.fgm3 = sum(T1_FGM3),
      sum_d.fga = sum(T1_FGA, T1_FGA3),
      sum_d.or = sum(T1_OR),
      sum_d.df = sum(T1_DR),
      sum_d.to = sum(T1_TO),
      sum_d.ftm = sum(T1_FTM),
      sum_d.fta = sum(T1_FTA)      
  )
  
  rbind(df1,df2) %>% 
    group_by(Season) %>% 
    summarize_all(sum)
}


ff_df2 = data.frame(TeamName = rep(MM21_teams, 18)) %>% 
  mutate(
    Season = rep(seq(2003, 2020, 1), 68)
  )

```

